Leadership across cultures and contexts: an exploratory analysis

### 2020-04-29 leadership_across_cultures_contexts.Rmd
Working on intro

<!-- INTRO TO MOTIVATE PAPER -->
Across human societies, the influence and agendas of leaders stands to substantially shape group structure and group outcomes. The structure of groups, their goals, and the demands of leaders can favor particular leadership qualities as well as the costs and benefits to individuals involved. Cultural norms, capacities of competing leaders, inter-group dynamics, and follower preferences can also impact leader agendas. The interactions between leader strategies, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural and biological evolutionary processes [@henrich_big_2015].

@@

<!-- WEIRD PEOPLE PROBLEMS -->
There is now well established consensus within behavioral and evolutionary human sciences on the importance of data from the full range of human cultural diversity and systemic biases in much of social science are well known [@henrich_weirdest_2010, *The weirdest people in the world?*]. @henrich_weirdest_2010 in their initial critique of the over-realiance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies, but also identified a contrast that, "compares people from modern industrialized societies with those from small-scale societies" (p.62), which has become the most salient interpretation of the paper and call-to-action. Since the seminal WEIRD people paper many social scientists have indeed taken up the challenge to diversify their samples and have sought out "non-WEIRD" participants and databases. However, categorizing populations as WEIRD or non-WEIRD and attempting to charachterize "non-WEIRD" social dynamics to learn about human psychology is a reductionist over-correction and arguably only a marginal scientific improvement from exclusively studying WEIRD populations.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and traditional societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. But such approaches are also at risk of failing to incorporate the message of the larger WEIRD people problem: diversity of cultural, social, and ecological contexts can be strong divers of human behavior and group dynamics.

>>>>

<!-- INTRO TO MOTIVATE PAPER -->

Despite decades of research and theorizing about leadership in numerous social science disciplines, to our knowledge there is no comprehensive or systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of cultures. Moreover, most research on leadership is based on data from Western or Westernized societies.

<!-- WEIRD PEOPLE PROBLEMS -->
For well over a century, anthropology has systematically investigated nearly every extant culture. Research on leadership within anthropology dates to the inception of the discipline. Early on, anthropologists (and some sociologists) realized that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; ]

Firth, 1927; Morgan, 1877; Myres, 1917]. Anthropologists documented some sort of leadership among every ethnographically studied culture [@brown_human_1991; @lewis_leaders_1974], and in many social contexts, such as families and kin groups (Dussart, 2000), ritual [@singh_cultural_2017], work groups [@macfarlan_competitive_2012], and conflicts between groups [@glowacki_evolutionary_2017]. For review, see @garfield_review_2019.

There is now well established consensus within behavioral and evolutionary human sciences on the importance of data from the full range of human cultural diversity and systemic biases in much of social science are well known [@henrich_weirdest_2010, *The weirdest people in the world?*]. @henrich_weirdest_2010 in their initial critique of the over-realiance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies, but also identified a contrast that, "compares people from modern industrialized societies with those from small-scale societies" (p.62), which has become the most salient interpretation of the paper and call-to-action. Since the seminal WEIRD people paper many social scientists have indeed taken up the challenge to diversify their samples and have sought out "non-WEIRD" participants and databases. However, categorizing populations as WEIRD or non-WEIRD and attempting to charachterize "non-WEIRD" social dynamics to learn about human psychology is a reductionist over-correction and arguably only a marginal scientific improvement from exclusively studying WEIRD populations.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and traditional societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. But such approaches are also at risk of failing to incorporate the message of the larger WEIRD people problem: diversity of cultural, social, and ecological contexts can be strong divers of human behavior and group dynamics.

Leaders substantially shape group structures and outcomes across cultures. The structure of groups and their goals, in turn, as well as their subsistence strategies, can favor particular leader qualities.

as well as the costs and benefits to individuals involved. Cultural norms, capacities of competing leaders, inter-group dynamics, and follower preferences can also impact leader agendas. The interactions between leader strategies, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural and biological evolutionary processes [@henrich_big_2015].

@@
